---
title: "Lecture 4 in-class note"
author: "Lau Wai Fung Raymond"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 0. Load the `tidyverse` package

This section loads the packages we need in this lecture.

```{r}
library(tidyverse)
```

### 1. Import and Export V-dem data

This section loads the VDEM dataset and describe its basic information.

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

### 2. Select economic development indicators

First, we look at the identifiers of the data.

```{r}
names(d)
```

We may use some alternative functions that provides information about the dataset. The `str()` provides not only variable names, but also their data types and a few example data points.

```{r}
str(d)
```

Usually, the second step of my data inquiry is having an overview of the *identifiers* of data points. In our case, the identifiers are country names, country IDs, and years. Using the `distinct()` function can effectively identify the distinct levels of *identifiers*

```{r}
d |> select(country_name, country_id, year) |>
  distinct()
```

```{r}
# Which countries are in this dataset
d |> select(country_name) |> distinct()
```

```{r}
d |> select(year) |> distinct()
```

Select both the country identifiers, GDP, and GDP per capita.

```{r}
d_gdp <- d |> 
  select(country_name, country_id, year, e_gdp, e_gdppc)

d_gdp
```

### Rename Column to Make Names Informative

```{r}
# d_gdp |> 
#  rename("GDP" = "e_gdp", "GDP_per_capita" = "e_gdppc", 
#         "Country" ="country_name", "ID" = "country_id", 
#         "Year" = "year")


d_gdp <- d_gdp |> 
  rename("GDP" = "e_gdp", "GDP_per_capita" = "e_gdppc", 
         "Country" ="country_name", "ID" = "country_id", 
         "Year" = "year")

d_gdp
```

### 4. `slice` Rows

```{r}
# Want country-years with highest GDP
d_gdp |> 
  slice_max(order_by = GDP, n = 10)
```

```{r}
# Get countries-years with the lowest GDP
d_gdp |> 
  slice_min(order_by = GDP, n = 10)
```

```{r}
set.seed(52)  #required to make research reproducible
d_gdp |> slice_sample(n = 10)  # Sample 10 observations
```

```{r}
d_gdp |> slice_sample(prop = 0.1)
```

### 5. Subset data by row

```{r}
# Want: 2000-2005 data
d_gdp |> 
  filter(Year >= 2000 & Year <= 2005)
```

```{r}
# store the new dataset that include China GDP
d_gdp_china <- d_gdp |> filter(Country == 'China')
```

```{r}
# Want: 2000-2005 data from China
d_gdp |> 
  filter(Year >= 2000 & Year <= 2005) |> 
  filter(Country == "China")
```

### 6. Arrange

```{r}
# Want: sort the data by GDP per capita
d_gdp |> arrange(GDP_per_capita)
```

```{r}
d_gdp |> arrange(-GDP_per_capita)
```

### 7. Perform (4) (5) (6) group by group: `group_by`, `ungroup`

**Task:** Create a dataset that contains the year of the higest development level for each country/ region respectively.

1.  Perform a data availability/ integrity check. Then aggregate the data into a new country-level dataset which contains the following indicators:
    1.  Average development level from 1984 to 2022.

    2.  Magnitude of growth from 1984 to 2022.

```{r}
# Want: For each country, we want year with the highest GDP
d_gdp |>
  group_by(Country) |>
  slice_max(order_by = GDP, n = 1)
```

#### Perform a data availability/ integrity check

```{r}
# How many entries are there for each country
d_gdp |>
  group_by(Country) |>
  count()
# ?count
```

```{r}
# Want: For each country, get the year when it has the worst GDP
d_gdp |>
  group_by(Country) |>
  slice_min(order_by = GDP, n = 1)
```

### 8. Create new columns in the data: `group_by`, `mutate`, `ungroup`

```{r}
# simple command using mutate
d_gdp |> mutate(New = 1)


d_gdp |> mutate(New = GDP)

d_gdp |> mutate(New = log(GDP))
d_gdp |> mutate(New = log(GDP)+1)

# Want: New column to be GDP relative to average GDP in the world 1984-2022
d_gdp |> mutate(GDP_over_avg = GDP/ mean(GDP, na.rm = TRUE))

# Want: New column to be GDP relative to average GDP of the country in the world 1984-2022
d_gdp |>
  group_by(Country) |>
  mutate(GDP_over_avg = GDP/ mean(GDP, na.rm = TRUE))
```

**Task:** Add the following economic indicators to the data:

1.  Country-year development level with reference to that of 1984.

2.  Year-on-year economic growth.

```{r}
# Country-year development level with reference to that of 1984.
d_gdp |> 
  group_by(Country) |>
  arrange(Year) |>
  mutate(GDP_over_1984 = GDP/first(GDP)) |>
  ungroup() |>
  arrange(Country, Year)
# first()
```

```{r}
# Year-on-year economic growth
# ?lag


d_gdp |>
  group_by(Country) |>
  arrange(Year) |>
  mutate(GDP_yoy_change = GDP - lag(GDP, n = 1)) |>
  ungroup() |>
  arrange(Country, Year)

# mutate function only help adding new columns, not for number of rows
```

### 9. Summarize the data: `group_by`, `summarise`, `ungroup`

```{r}
# Want: Average GDP level of the world
d_gdp |> summarise(gdp_average = mean(GDP, na.rm = TRUE),
                   gdp_per_capita_average = mean(GDP_per_capita, na.rm = TRUE))
```

**Task:** Perform a data availability/ integrity check. Then aggregate the data into a new country-level dataset which contains the following indicators:

1.  Average development level from 1984 to 2022.

2.  Magnitude of growth from 1984 to 2022.

```{r}
# Data availability/ integrity check
d_gdp |>
  # Create a column that indicates whether the value is missing
  mutate(GDP_missing = as.numeric(is.na(GDP)), .after = GDP) |>
  group_by(Country) |>
  summarise(N_GDP_missing = sum(GDP_missing))

# ?is.numeric
```

```{r}
# Average development level
d_gdp |>
  group_by(Country) |>
  summarise(GDP_average = mean(GDP, na.rm = TRUE),
            GDPpc_average = mean(GDP_per_capita, na.rm = TRUE))

```

```{r}
# GDP growth and GDP per capita growth: comparing 2019 with 1984
d_gdp |>
  filter(Year >= 1984 & Year <= 2019) |>
  group_by(Country) |>
  arrange(Year) |>
  summarise(GDP_growth_2020_1984 = (last(GDP) - first(GDP)) / first(GDP),
            GDPpc_growth_2019_1984 = (last(GDP_per_capita) - first(GDP_per_capita))
            /first(GDP_per_capita)) |>
  ungroup() |>
  arrange(Country)
```

### Details of Pipe operation notes refers to Lecture_4_note_rmd

#### Why piping?

Pipe is useful when you are conducting a series of operation on your data but want to minimize the number of intermediate outputs produced.

```{r}
# STEP 1: Subset variables
d_gdp <- d |> select(country_name, country_id, year, e_gdp, e_gdppc)

# STEP 2: Rename variables
d_gdp_renamed <- d_gdp |>
  rename("GDP" = "e_gdp", "GDP_per_capita" = "e_gdppc",
         "Country" = "country_name", "ID" = "country_id",
         "Year" = "year")

# STEP 3: Filter down to China
d_gdp_china <- d_gdp_renamed |> filter(Country == "China")

# STEP 4: Filter down to 2000 - 2005
d_gdp_china_2000_2005 <- d_gdp_china |> filter(Year >= 2000 & Year <= 2005)

d_gdp_china_2000_2005
```
