---
title: "POLI3148 Lecture 4 in class exercise 2"
author: "Lau Wai Fung Raymond"
date: "2023-10-05"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load the `tidyverse` package

```{r}
library(tidyverse)
```

#### Import and Export V-dem data

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

### 1. Codebook look up

**i. What indicators regarding the quality of education are available in the V-Dem datasets?**

In the V-Dem dataset, there is two indicators that relates to the quality of education, which is Educational inequality (e_peedgini), and Education 15+ (e_peaveduc).

**ii. What are the data's coverage (i.e., for which countries and years do we have data?)**

```{r}
# Find out which country and year are available on the dataset
d |> select(country_name, country_id, year) |> distinct()
```

Based on the codebook, the data coverage (year) for Education 15+ indicator is available from 1820 to 2022, while the data coverage (year) for Educational inequality indicator is available form 1850 to 2010.

**iii. What are their sources? Provide the link to least 1 source.**

The V-Dem data is download from [V-Dem Dataset](https://v-dem.net/) website.

```{r}
# Load the `tidyverse` package before importing the dataset
library(tidyverse)
```

```{r}
# Import the V-Dem dataset that used for this in-class exercise 2
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

### 2. Subset by columns

**i. Create a dataset containing only the country-year identifiers and indicators of education quality.**

```{r}
# create a new dataset that contains country-year identifiers, and education quality indicators

d_edu <- d |>
  select(country_name, country_id, year, e_peaveduc, e_peedgini)

d_edu
```

**ii. Rename the columns of education quality to make them informative.**

```{r}
# Rename the columns of educational quality indicators
d_edu <- d_edu |> 
  rename("avg_year_education" = "e_peaveduc", "education_inequality" ="e_peedgini",
         "Country" = "country_name", "ID" = "country_id",
         "Year" = "year")

d_edu
```

### 3. Subset by rows

**i.** **List 5 countries-years that have the highest education level among its population.**

```{r}
# List five countries-years with highest education level among the population
d_edu |> slice_max(order_by = avg_year_education, n = 5)
```

**ii.** **List 5 countries-years that suffer from the most severe inequality in education.**

```{r}
# Get 5 country-years that have the most severe educational inequality
d_edu |> slice_min(order_by = education_inequality, n = 5)
```

### 4. Summarise the data

**i.** **Check data availability: For which countries and years are the indicators of education quality available**

```{r}
# Data integrity check
# first step: check whether the educational quality are available in different countries and year
d_edu |>
  mutate(avg_year_education_missing = as.numeric(is.na(avg_year_education)), 
         .after = avg_year_education) |>
  mutate(education_inequality_missing = as.numeric(is.na(education_inequality)),
         .after = education_inequality) |>
  arrange(Country)
```

```{r}
# second step
# For each country, calculate the missing data in education quality 
d_edu |>
    mutate(avg_year_education_missing = as.numeric(is.na(avg_year_education)), 
           .after = avg_year_education) |>
    mutate(education_inequality_missing = as.numeric(is.na(education_inequality)), 
           .after = education_inequality) |>
    group_by(Country) |>
    summarise(N_avg_years_education_missing = sum(avg_year_education_missing),
              N_education_inequality_missing = sum(education_inequality_missing))
```

**ii.** **Create two types of country-level indicators of education quality**

```{r}
# Average level of education quality from 1984 to 2022
d_edu |> 
  group_by(Country) |>
  summarise(average_education_years = mean(avg_year_education, na.rm = TRUE),
            average_education_inequality = mean(education_inequality, na.rm = TRUE))
```

```{r}
# Question: change of education quality from 1984 to 2022
# Due to data availability issue, comparing 2010 with 1984 
d_edu |>
  filter(Year >= 1984 & Year <= 2010) |> 
  group_by(Country) |>
  arrange(Year) |>
  summarise(educationyears_2010_1984 = (last(avg_year_education) - first(avg_year_education))/ 
              first(avg_year_education),
            education_inequality_2010_1984 = (last(education_inequality) - first(education_inequality))
            / first(education_inequality)) |>
  ungroup() |>
  arrange(Country)
```

**iii.** **Examine the data and briefly discuss: Which countries perform the best and the worst in terms of education quality in the past four decades?**

Comparing the country's performance in education, Germany has been the country that performs the best in the average years of education in the past decades, citizens older than 15 had received 12.939 years of education on average. On the other hand, Burkina Faso represents a country with the lowest average years of education. On average, less than a year of education for citizens over 15 years old (value = 0.982).

Likewise, Burkina Faso is the country that has suffered the most severe educational inequality on average, compared to other countries. The Gini coefficient of unequal education in Burkina Faso is 91.26. In contrast, Austria is the country that had the lowest average educational inequality, whereas their Gini coefficient only consists of 6.35.
