---
title: "Lecture 5 in-class notes"
author: "Lau Wai Fung Raymond"
date: "2023-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## In-Class Demo

```{r}
library(tidyverse)
```

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

## Example: The V-Dem Data

Focus on the economic indicators: GDP and GDP per capita.

```{r}
d_gdp <- d |> 
  select(country_text_id, year, e_gdp, e_gdppc) |>
  rename("gdp" = "e_gdp", "gdppc" = "e_gdppc")

d_gdp |> print(n = 3)
```

# Reshape a Table

## Wide to Long: `pivot_longer`

```{r}
d_gdp_long <- d_gdp |>
  pivot_longer(cols = c("gdp", "gdppc"), 
               names_to = "variable", values_to = "value")

d_gdp_long |> print(n = 4)
```

## Long to Wide: `pivot_wider`

Task: Reverse the above `pivot_long` operation. \# turn the dataset back to the original shape

```{r}
d_gdp_wide_1 <- d_gdp_long |>
  pivot_wider(names_from = "variable", values_from = "value")

d_gdp_wide_1 |> print(n = 4)
```

## Long to Wide: `pivot_wider`

Task: Make `year` the column variable.

```{r}
d_gdp_wide_2 <- d_gdp_long |>
  pivot_wider(names_from = "year", values_from = "value")

d_gdp_wide_2 |> print(n = 2)
```

## Long to Wide: `pivot_wider`

Task: Make `country_text_id` the column variable.

```{r}
d_gdp_wide_3 <- d_gdp_long |>
  pivot_wider(names_from = "country_text_id", values_from = "value")

d_gdp_wide_3 |> print(n = 2)
```

## Why Do We Reshape Data?

-   **For data cleaning:** Sometime it is much easier to clean the data after reshaping

-   **For data visualization:** Some data visualization functions only take tables shaped in a specific way

-   **For data sharing:** Sometimes you want to export the data for human readers (e.g., data coding/ labeling)

*"But I am sure Excel can do the same thing!" It can do it for HUGE data reliably and fast. And the process is replicable.*

### 1. Reshape the V-Dem dataset

```{r}
# Want: Each row contain country-year-variable

d |> select(country_name) |> distinct()

d_subset <- d |> select(country_name, year, starts_with("e_"))

d_subset_long <- 
  d_subset |> pivot_longer(cols = starts_with("e_"))
# each of the columns transforms as in own row, 
```

```{r}
# each rows become a columns i.e. year
d_subset_wide_year <-
  d_subset_long |> pivot_wider(names_from = year, values_from = value)
```

### 2. Stack multiple subsets of the V-Dem datasets

```{r}
d_VdemIndex <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_index.csv")
```

```{r}
d_stack <- bind_cols(d, d_VdemIndex)
```

```{r}
# Want: Stack two tables by rows?
d_1945_1983 <- read_csv("_DataPublic_/vdem/1945_1983/vdem_1945_1983_external.csv")
```

```{r}
d_1945_2022 <- bind_rows(d, d_1945_1983)

d_1945_2022 |> select(year) |> distinct() |> arrange(year)
```
