---
title: "Lecture 7 in class notes (second-half)"
author: "Lau Wai Fung Raymond"
date: "2023-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Housekeeping

## Load Data

```{r}
library(tidyverse)
theme_set(theme_bw()) # Set my default theme for the whole document

d <- readRDS("Lecture_7/data/wealth_and_health.rds")
d |> print(n = 3)
```

## `ggplot` syntax

-   Consider: "Painting" your data on a canvas, layer by layer

-   Steps

    -   Specify the data you want to viaulize

    -   Specify the type(s) of visualization tools

    -   Label the figure (titles of the figure, axes, and legends)

-   Command: Use **`+`** to connect layers (ATTENTION: not `|>` )

Note: [***Always***]{.underline} properly label your figures!

## `ggplot` syntax: example

```{r}
{r, fig.height=3, fig.width=9, warning=FALSE}
d |> # Specify the data frame to plot
  ggplot(aes(x = gdppc, y = life_expectancy)) + # Specify the variables to plot
  geom_point() + # Choose the visualization tool: point
  labs(x = "GDP per capita", y = "Life Expectancy", # set up the axes' titles
       title = "Wealth and Health in the World (1800-2019)", # set up the figure's title
       caption = "By Lau Wai Fung. Data source: V-Dem v.13") # set up footnote
```

## Variable Types

Visualization tools to use largely depends on variable types

-   "Quantitative" (Continuous, Count)

    -   GDP per capita

    -   Life expectancy

    -   Population

    -   Infant mortality

-   Categorical

    -   Binary: Binary "democracy" indicator

    -   Nominal: Region

    -   Ordinal: Lexical Index of Electoral Democracy

# Quant. X 1

## Viaulize One Quantitative Variable

-   Summary Statistics / exploratory statistics

    -   Min, max, median, mean, standard deviation, variance,

    -   Quantiles, quintiles

-   Visualization

    -   Histogram

    -   Density Plot

-   Advanced

    -   Skewed distribution (may hard to interpret the numbers when log down the original numbers)

    -   Annotate cases (put label in case of interst in project)

## Summary Statistics

```{r}
summary(d$gdppc) # Quick summary
length(d$gdppc) # Number of observations
is.na(d$gdppc) |> sum() # Number of missing values
```

## Summary Statistics (con'd)

```{r}
mean(d$gdppc, na.rm = TRUE) # Mean
var(d$gdppc, na.rm = TRUE) # Variance
sd(d$gdppc, na.rm = TRUE) # Standard deviation
quantile(d$gdppc, na.rm = TRUE) # Quantile
quantile(d$gdppc, seq(0, 1, 0.1), na.rm = TRUE) # Quintile
```

## Histogram

```{r, fig.width=9, fig.height=3}
d |> 
  ggplot(aes(x = gdppc)) +
  geom_histogram() + # Draw a histogram
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

## Histogram (customized)

```{r, fig.width=10, fig.height=2, warning=FALSE}
d |> ggplot(aes(x = gdppc)) + geom_histogram(bins = 10) +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

```{r, fig.width=10, fig.height=2, warning=FALSE}
d |> ggplot(aes(x = gdppc)) + geom_histogram(bins = 50, fill = "blue", color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

## Density

```{r, fig.width=9, fig.height=3, warning=FALSE}
d |> 
  ggplot(aes(x = gdppc)) +
  geom_density() + # Draw a density plot
  labs(x = "GDP per capita", y = "Density", title = "Density of GDP per capita")
```

## Density (customized)

```{r, fig.width=9, fig.height=3, warning=FALSE}
d |> 
  ggplot(aes(x = gdppc)) +
  geom_density(fill = "gray", color = "blue") + # Draw a density plot
  labs(x = "GDP per capita", y = "Density", title = "Density of GDP per capita")
```

## Skewed Distribution

-   **Challenge:** Quantitative data may have skewed distributions, which can make the visualization uninformative.

-   **Solution:**

    -   **Re-scale the axes**

    -   **Transform the variable(s)**

## Skewed Distribution: Re-scale Axes

```{r, fig.width=10, fig.height=2, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc)) + geom_histogram() +
  scale_x_sqrt() + # Scale the x axis to make distribution easy to read
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

```{r, fig.width=10, fig.height=2, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc)) + geom_histogram() +
  scale_x_log10() + # Scale the x axis to make distribution easy to read
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita",
       caption = "X-axis is re-scaled.")
```

## Skewed Distribution: Transform Variables

Alternatively you can first transform the variable before your visualization

```{r, fig.width=10, fig.height=2, warning=FALSE, message=FALSE}
d |> mutate(gdppc_sqrt = sqrt(gdppc)) |> # Transform the variable before visualization
  ggplot(aes(x = gdppc_sqrt)) + geom_histogram() +
  labs(x = "sqrt(GDP per capita)", y = "Count", title = "Histogram of GDP per capita")
```

```{r, fig.width=10, fig.height=2, warning=FALSE, message=FALSE}
d |> mutate(gdppc_log = log(gdppc)) |> # Transform the variable before visualization
  ggplot(aes(x = gdppc_log)) + geom_histogram() +
  labs(x = "log(GDP per capita)", y = "Count", title = "Histogram of GDP per capita")
```

## Skewed Distribution: Note

```{r}
summary(d$gdppc)
```

**The method only work when your variable contains NON-NEGATIVE values only.**

## Annotate Case(s) of Interest

```{r, fig.width=9, fig.height=2, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc)) + geom_histogram() +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

## Annotate Case(s) of Interest (con'd)

```{r, fig.width=9, fig.height=2, warning=FALSE, message=FALSE}
d_anno <- d |> filter(country_text_id == "CHN", year == 2019) |> mutate(anno = paste(country_text_id, year), .after = year)
print(d_anno)
# A "manual" approach
d |> ggplot(aes(x = gdppc)) + geom_histogram() + 
  geom_vline(aes(xintercept = 15.4), linetype = "dashed", color = "red") +
  annotate("text", x = 15.4, y = 7500, hjust = 0, label = "China 2019", color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
# How do I get y = 7500? Observe the y-axis in the previous figure.
```

## Annotate Case(s) of Interest (con'd)

```{r, fig.width=9, fig.height=2, warning=FALSE, message=FALSE}
# An automated approach
d |> ggplot(aes(x = gdppc)) + geom_histogram() + 
  geom_vline(data = d_anno, aes(xintercept = gdppc), linetype = "dashed", color = "red") +
  geom_text(data = d_anno, aes(x = gdppc, y = 7500, label = anno), hjust = 0, color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

Why bother with the automated approach? It helps when you want to annotate multiple cases.

## Annotate Case(s) of Interest (con'd)

```{r, fig.width=9, fig.height=2, warning=FALSE, message=FALSE}
d_anno <- d |> filter(country_text_id %in% c("CHN", "JPN", "USA"), year == 2019) |>   mutate(anno = paste(country_text_id, year), .after = year)
print(d_anno)

d |> ggplot(aes(x = gdppc)) + geom_histogram() + 
  geom_vline(data = d_anno, aes(xintercept = gdppc), linetype = "dashed", color = "red") +
  geom_text(data = d_anno, aes(x = gdppc, y = 7500, label = anno), hjust = 0, color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

## Annotate Case(s) of Interest (con'd)

```{r, fig.width=9, fig.height=2, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc)) + 
  geom_density() + 
  geom_vline(data = d_anno, aes(xintercept = gdppc), linetype = "dashed", color = "red") +
  geom_text(data = d_anno, aes(x = gdppc, y = 0.15, label = anno), hjust = 0, color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
# Why do I change y to 0.15? Through observation again... cuz the unit is different
```

# Quant. X 2

## Visualize Two Quantitative Variables

-   Plots

    -   Scatter plot

    -   Scatter plot with a trend line

-   Advanced

    -   Skewed distribution

    -   Annotate data points

    -   2D "Histogram"

-   Summary Statistics: Correlation

## Scatter Plot

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## Scatter Plot (customized)

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, color = "navy", shape = 3, size = 0.5, stroke = 1) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
# alpha is the transparency of the dots
# search for R color
```

## Scatter Plot with a Trend Line

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_smooth() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## Scatter Plot with a Trend Line (customized aesthetic)

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_smooth(se = FALSE, color = "red") +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")

# se = standard error; the shadow area
```

## Scatter Plot with a Trend Line (customized model)

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_smooth(method = "lm") +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

**This is a BAD example, because the relationship between the two variables is clearly not linear!**

## Skewed Distribution: Re-scale Axes

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, color = "navy", shape = 3, size = 0.5, stroke = 1) +
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)", caption = "x-axis is rescaled")
```

## Skewed Distribution: Transform variables

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> 
  mutate(gdppc_log = log(gdppc)) |>
  ggplot(aes(x = gdppc_log, y = life_expectancy)) +
  geom_point(alpha = 0.3, color = "navy", shape = 3, size = 0.5, stroke = 1) +
  geom_smooth(method = "lm", color = "red") +
  labs(x = "log(GDP per capita)", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## Annotate Data Points: `geom_text`

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_text(data = d_anno, aes(x = gdppc, y = life_expectancy, label = anno), color = "red", size = 2.5) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## Annotate Data Points: `geom_label`

`geom_text` might not be able to highlight your data points enough (especially when your scatter plot has a lot of data points). `geom_label` may help.

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_label(data = d_anno, aes(x = gdppc, y = life_expectancy, label = anno), color = "red", fill = "wheat" , size = 2.5) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## Annotate MANY Data Points: Problem

`geom_label` and `geom_text` may get you uninformative annotation if the points you want to annotate cluster in a small area of your figure.

```{r, fig.width=12, fig.height=4, warning=FALSE, message=FALSE}
d_anno_2 <- d |> filter(year >= 2010, country_text_id == "CHN") |> 
  mutate(anno = paste(country_text_id, year), .after = year)

d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  # geom_point(data = d_anno_2, aes(x = gdppc, y = life_expectancy), color = "red", size = 0.5) +
  geom_label(data = d_anno_2, aes(x = gdppc, y = life_expectancy, label = anno), color = "red", size = 2)
```

## Annotate MANY Data Points: `geom_label_repel`

When you need to annotate a few points that may cluster in a small area of your figure, `geom_label_repel` can adjust the labels' locations to avoid overlaps.

```{r, fig.width=12, fig.height=4, warning=FALSE, message=FALSE}
library(ggrepel)
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  # geom_point(data = d_anno_2, aes(x = gdppc, y = life_expectancy), color = "red", size = 0.5) +
  geom_label_repel(data = d_anno_2, aes(x = gdppc, y = life_expectancy, label = anno), color = "red", size = 2, max.overlaps = 20) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")

# aware to avoid many overlaps point
# aware the error warning message
```

## 2D "Histogram"

When there have many data point in scatter plot (less informative), this 2D histogram will become useful

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_bin_2d(bins = 80) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## 2D "Histogram"

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_bin_2d(bins = 80) + scale_x_log10() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## 2D "Histogram"

If you find the default color scheme (dark to light blue) to no adequately visualize the different frequencies, you may use a different scheme that offers higher "contrast" between low and high values. `scale_fill_viridis_c` **is one of my favorite. Can add option "A", "B" etc after** scale_fill_viridis_c

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_bin_2d(bins = 80) + scale_x_log10() +
  scale_fill_viridis_c() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## 2D "Histogram"

```{r, fig.width=9, fig.height=3, warning=FALSE, message=FALSE}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_hex(bins = 80) + scale_x_log10() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## Summary Statistics: Pearson Correlation

Pearson correlation (aka. Pearson's *r*), ranging from -1 to 1, is our typical indicator of the correlation between two quantitative variables.

```{r}
cor(d$life_expectancy, d$gdppc, use = "complete.obs", method = "pearson")
cor.test(d$life_expectancy, d$gdppc, use = "complete.obs", method = "pearson")
```

## Summary Statistics: Pearson Correlation

To use Pearson correlation, the two variables should meet the following assumptions:

-   **Continuous**

-   **Linear relationship**

-   **No outlier**

-   **Normally distributed**

## Summary Statistics: Spearman Correlation

If the data seem to break at least one of the assumptions for Pearson correlation, you may use Spearman correlation (aka. Spearman's $\rho$). In spearman correlation, based on ranking of correlation instead of actual value

```{r}
cor(d$life_expectancy, d$gdppc, use = "complete.obs", method = "spearman")
cor.test(d$life_expectancy, d$gdppc, use = "complete.obs", method = "spearman")
```

# Quant. X 3

## Visualize 3 Quantitative Variables

-   Correlation Matrix

-   Bubble Plot: 3rd variable visualized with size

-   Make it colorful: 3rd variable visualized with color

## Correlation Matrix

```{r, fig.height=4, message=FALSE, warning=FALSE}
# install.packages("GGally")
library(GGally)
d |>
  select(gdppc, life_expectancy, population) |>
  ggpairs()
```

## Correlation Matrix

```{r, fig.height=4, message=FALSE, warning=FALSE}
d |> select(gdppc, life_expectancy, population) |>
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "Population"), # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")), # Upper triangle: Spearman correlation
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), # Diagnal: histogram
    lower = list(continuous = wrap("smooth", alpha = 0.1, color = "gray"))) # Lower triag: + linear trend line
```

## Correlation Matrix

```{r, fig.height=4, message=FALSE, warning=FALSE}
# barDiag = barplot
# smooth = smooth lines
d |> select(gdppc, life_expectancy, population) |>
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "Population"), 
    # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")), 
    # Upper triangle: Spearman correlation
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), # Diagnal: histogram
    lower = list(continuous = wrap("smooth_loess", alpha = 0.1, color = "gray"))) 
# Lower triag: + loess trend line
```

## Correlation Matrix

```{r, fig.height=2, message=FALSE, warning=FALSE}
summary(d$population) # The population variable is skewed. Consider transofmration
d |> select(gdppc, life_expectancy, population) |>
  mutate(population = log(population)) |> # log transformation
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "log(Population)"), # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")), # Upper triangle: Spearman correlation
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), # Diagnal: histogram
    lower = list(continuous = wrap("smooth_loess", alpha = 0.1, color = "gray"))) # Lower triag: + loess trend line
```

## Play with `ggpairs`

Check out its documentation: <https://www.rdocumentation.org/packages/GGally/versions/1.5.0/topics/ggpairs>

## Bubble Plot

Let population design the point size

```{r}
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population)) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (2019)", size = "Population") +
  theme(legend.position = "none")
```

## Bubble Plot (customized)

```{r, message=FALSE, warning=FALSE}
# alpha = transprancy (lower then more transparent)
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population), alpha = 0.3) +
  labs(x = "GDP per capita", y = "Life Expectancy", size = "Population",
       title = "Wealth and Health in the World (2019)")
```

## Bubble Plot (annotated)

```{r, message=FALSE, warning=FALSE}
d_anno_pop <- d |> filter(year == 2019) |> slice_max(order_by = population, n = 10)

# can refers to different dataset in geom_text syntax
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population), alpha = 0.3) +
  geom_text(data = d_anno_pop, aes(x = gdppc, y = life_expectancy, label = country_text_id, size = population)) +
  labs(x = "GDP per capita", y = "Life Expectancy", size = "Population",
       title = "Wealth and Health in the World (2019)",
       caption = "Note: World's top 10 most populated countries are annotated.") +
  theme(legend.position = "none") # Remove the legend
```

## Make it Colorful

Strengthen the visual message from the third variable

```{r, message=FALSE, warning=FALSE}
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population, color = population), alpha = 0.3) +
  labs(x = "GDP per capita", y = "Life Expectancy", 
       size = "Population", color = "Population", fill = "Population",
       title = "Wealth and Health in the World (2019)")
```

## Make it Colorful (customized)

```{r, message=FALSE, warning=FALSE}
# _c mean continuous
# _d for discrete data (categorical)
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population, color = population), alpha = 0.3) +
  scale_color_viridis_c(option = "A", trans = "log", direction = -1) +
  labs(x = "GDP per capita", y = "Life Expectancy", 
       size = "Population", color = "Population", fill = "Population",
       title = "Wealth and Health in the World (2019)")
```

# Quant. X 4+?

## Warning

It is *technically* feasible to plot more than 3 variables in the same figure, but...

-   You don't want your figure to be too "busy"

-   "Less is more"

## Add infant mortality to the visualization

In this case, using color to visualize infant mortality is reasonable.

```{r, message=FALSE, warning=FALSE}
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population, color = infant_mortality), alpha = 0.3) +
  scale_color_viridis_c(option = "B", direction = -1) +
  labs(x = "GDP per capita", y = "Life Expectancy", 
       size = "Population", color = "Infant Mortality",
       title = "Wealth and Health in the World (2019)") +
  theme_classic()
```

## Correlation Matrix

```{r, message=FALSE, warning=FALSE}
# library(GGally)
d |> filter(year == 2019) |>select(gdppc, life_expectancy, population, infant_mortality) |>
  mutate(population = log(population)) |>
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "log(Population)", "Infant Mortality"), # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")),
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), 
    lower = list(continuous = wrap("smooth_loess", alpha = 0.3, color = "gray")))
```
